### 2020-05-01 final_report/conclusion.tex

After a thorough review of the literature, we developed a novel adaptive
sampling algorithm, QASS, capable of interfacing with any of the studied
surrogates. Preliminary testing on a toy theory, qualitatively comparable to
the MC TBR model, demonstrated the effectiveness of QASS and behavioural trends
consistent with the design of the algorithm. \textit{[Insert numerical results % TODO
for QASS.]} Further optimisation over the hyperparameter space has strong
potential to increase this performance, allowing for future deployment of QASS
in coalition with any of the most effective identified surrogates to facilitate
learning during evaluation of the MC TBR model.

>>>>

After a thorough review of the literature, we developed a novel adaptive
sampling algorithm, QASS, capable of interfacing with any of the studied
surrogates. Preliminary testing on a toy theory, qualitatively comparable to
the MC TBR model, demonstrated the effectiveness of QASS and behavioural trends
consistent with the design of the algorithm. With~\num{100000} initial samples and 100 incremental samples per iteration, QASS achieved a ${\sim}60\%$ decrease in surrogate error as compared with a baseline random sampling scheme. Equivalently, the same surrogate error as the baseline could be achieved with ${\sim}6\%$ fewer total samples of the expensive theory. Further optimisation over the hyperparameter space has strong
potential to increase this performance, in particular by decreasing the required quantity of initial samples. This will allow for future deployment of QASS
in coalition with any of the most effective identified surrogates to facilitate
learning during evaluation of the MC TBR model.

### 2020-05-01 final_report/data.tex

The MC TBR model that we understand to be the expensive function for our
surrogate modelling is fundamentally a Monte Carlo simulation based on the
OpenMC framework~\cite{ROMANO201590}. As input the software expects a set of 19~parameters, discrete and
continuous, that are fully listed in~\cref{tbl:params}. Following a brief period of
time (usually of the order of tens of seconds), during which a fixed number of
events is generated, the simulation outputs the
mean and the standard deviation of the TBR aggregated over the simulated run. The
former of these two we accept to be the output TBR value that is subject
to approximation.

@@

In our work, we often reference TBR points or samples. These are simply vectors
in the feature space generated by Cartesian product of domains of all
features---parameters from~\cref{tbl:params}.

@@

While we deliberately make no assumptions about the internal properties of the
MC TBR simulation, effectively treating it as a black box model, our means of
studying its behaviour are limited to inspection of its outputs at various
points in the feature space, and inference thereof. We thus require
sufficiently large and representative quantities of samples to ensure that
surrogates can be trained to approximate the MC TBR model.

With grid search in such a high-dimensional domain clearly intractable, we
selected uniform pseudo-random sampling\footnote{Continuous and discrete
parameters are drawn from a corresponding uniform distribution over their
domain, as defined in~\cref{tbl:params}. For reproducibility, each run uses a seed equal to its number.} to generate large amounts of feature
configurations that we consider to be independent and unbiased. For evaluation
of the expensive MC TBR model, we utilise parallelisation offered by
the HPC infrastructure available at UCL computing facilities. To this end, we
designed and implemented the Approximate TBR Evaluator---a Python software package capable of
sequential evaluation of the multi-threaded OpenMC simulation on batches of
previously generated points in the feature space.
We deployed ATE at the UCL Hypatia cluster RCIF partition, which consists of
4~homogeneous nodes, each containing 40 cores.\footnote{Hypatia RCIF nodes use
40~Intel\textsuperscript{\textregistered} Xeon\textsuperscript{\textregistered}
Gold 6148 CPUs with frequency 2.40~GHz and 376 GB RAM.} We completed three data
generation runs, which are summarised in~\cref{tbl:sampling-runs}.

@@

Model training over high-dimensional parameter spaces (illustrated
in~\cref{fig:tbr-vs-features}) may be improved in many
aspects by carefully reducing the number of variables used to describe the
space. For many applications, feature selection strategies succeed in
identifying a sufficiently representative subset of the original input
variables; however, all given variables were assumed to be physically relevant
to the MC TBR model. Feature extraction methods, on the other hand, aim to
identify a transformation of the parameter space which decreases dimensionality; even if no individual parameter is separable from the space, some linear combinations of parameters or nonlinear functions of parameters may be.

@@

Kriging is a geostatistical surrogate modelling technique which relies on
correlation functions over distance (lag) in the feature space \cite{Bouhlel2018}. Although kriging performed poorly for our use case due to high dimensionality, these correlation measures gave insight into similarities between discrete-parameter slices of the data.

\Cref{fig:var} shows the Matheron semivariance \cite{Matheron1963} for three discrete slices with coolant material varied, but all other discrete parameters fixed. Fits \cite{KrigingFig} to the Matérn covariance model confirmed numerically that the coolant material is the discrete parameter with the greatest distinguishability in the MC TBR model.

@@

In particular, we consider two equally-sized sets\footnote{Each set contained
\num{100000}~samples from batches 0-99 of the corresponding runs.} of samples:
(i)~a subset of data obtained
from run~1 and (ii)~a subset of a single slice obtained from run~2. Our
expectation was that while the former case would provide meaningful insights into
correlations within the feature space, the latter would validate our
autoencoder implementation by analysing a set of points that are trivially
reducible in dimensionality due to a number of fixed discrete features.

>>>>

The expensive MC TBR model is fundamentally a Monte Carlo simulation based on the
OpenMC framework~\cite{ROMANO201590}. As input the software expects 18~parameters, discrete and
continuous, that are fully listed in~\cref{tbl:params}. During evaluation, which
usually takes units of seconds, a fixed number of events is simulated, and the
results are given in terms of the mean and the standard deviation of the~TBR aggregated
over the simulated run. The former of these two we accept to be the output TBR
value that is subject to approximation.

@@

In the following sections, we often reference TBR points or samples. These are simply vectors
in the feature space generated by Cartesian product of domains of all
features---parameters from~\cref{tbl:params}.

@@

In our work, we deliberately make no assumptions about the internal properties of the
MC TBR simulation, effectively treating it as a black box model. This limits our
means of studying its behaviour to inspection of its outputs at various
points in the feature space. We therefore require
sufficiently large and representative quantities of samples to ensure that
surrogates can be trained to approximate the MC TBR model.

With a grid search in such a high-dimensional domain clearly intractable, we
selected uniform pseudo-random sampling\footnote{Continuous and discrete
parameters are drawn from a corresponding uniform distribution over their
domain, as defined in~\cref{tbl:params}. For reproducibility, each run uses a seed equal to its number.} to generate large amounts of feature
configurations that we consider to be independent and unbiased. For evaluation
of the expensive MC TBR model, we utilise parallelisation offered by
the HPC infrastructure available at UCL computing facilities. To this end, we
designed and implemented the Approximate TBR Evaluator---a Python software package capable of
sequential evaluation of the multi-threaded OpenMC simulation on batches of
previously generated points in the feature space.
Having deployed ATE at the UCL Hypatia cluster,\footnote{The Hypatia RCIF
	partition is comprised of 4~homogeneous nodes. Each node is installed with
	376~GB RAM and 40~Intel\textsuperscript{\textregistered}
	Xeon\textsuperscript{\textregistered} Gold 6148 CPUs of clock frequency
	2.40~GHz.} we completed 3~data
generation runs that are summarised in~\cref{tbl:sampling-runs}.

@@

Model training over high-dimensional parameter spaces (illustrated
in~\cref{fig:tbr-vs-features}) may be improved in many
aspects by carefully reducing the number of variables used to describe the
space. For many applications, feature selection strategies succeed in
identifying a sufficiently representative subset of the original input
variables. However, all given variables were assumed to be physically relevant
to the MC TBR model. On the other hand, feature extraction methods aim to
identify a transformation of the parameter space which decreases dimensionality; even if no individual parameter is separable from the space, some linear combinations of parameters or nonlinear functions of parameters may be.

@@

Kriging is a geostatistical surrogate modelling technique which relies on
correlation functions over distance (lag) in the feature space~\cite{Bouhlel2018}. Although kriging performed poorly for our use case due to high dimensionality, these correlation measures gave insight into similarities between discrete-parameter slices of the data.

\Cref{fig:var} shows the Matheron semivariance~\cite{Matheron1963} for three discrete slices with coolant material varied, but all other discrete parameters fixed. Fits~\cite{KrigingFig} to the Matérn covariance model confirmed numerically that the coolant material is the discrete parameter with the greatest distinguishability in the MC TBR model.

@@

In particular, we consider two equally-sized sets\footnote{Each set contained
\num{100000}~samples from batches 0-99 of the corresponding runs.} of samples:
(a)~a subset of data obtained
from run~1 and (B)~A SUBSet of a single slice obtained from run~2. Our
expectation was that while the former case would provide meaningful insights into
correlations within the feature space, the latter would validate our
autoencoder implementation by analysing a set of points that are trivially
reducible in dimensionality due to a number of fixed discrete features.

### 2020-05-01 final_report/results.tex

An increase in MCMC candidate samples was seen to have a positive but very weak
effect on final surrogate precision, suggesting that the runtime of MCMC on each
iteration can be limited for increased efficiency. -- Awaiting test results on % TODO
initial sample quantity --. The most complex dynamics arose with the adjustment
of sample increment, shown in~\cref{fig:qassincr}. For each tested initial
sample quantity N, the optimal number of step samples was seen to be
well-approximated by $\sqrt{N}$. The plotted error trends suggest that
incremental samples larger than this optimum give slower model improvement on
both the training and evaluation sets, and a larger minimum error on the
evaluation set. This performance distinction is predicted to be even more
significant when trained on the expensive MC model, where the number of sample
evaluations will serve as the primary bottleneck for computation time.
\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=1.1\linewidth]{fig6a_qassincrsamp.pdf}
    \end{subfigure}%
    \hfill
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=1.1\linewidth]{fig6b_qassincrtime.pdf}
    \end{subfigure}
    \caption{QASS absolute training error over total sample quantity (left) and number of iterations (right). MAE represents surrogate error on the adaptively-sampled training/test set, and E\_MAE on the independent evaluation sets.}
    \label{fig:qassincr}
\end{figure}

@@

Such tests revealed that while QASS has unmatched performance on its own
adaptively-sampled training set, it is outperformed by the baseline scheme on
uniformly-random evaluation sets. We suspected that while QASS excels in
learning the most strongly peaked regions of the TBR theory, this comes at the
expense of precision in broader, smoother regions where uniformly random
sampling suffices. Therefore a mixed scheme was implemented, with half MCMC
samples and half uniformly random samples incremented on each iteration, which
is also shown in~\cref{fig:qasssampling}.

>>>>

An increase in initial samples with increment held constant had a strong impact on final surrogate precision, an early confirmation of basic functionality. An increase in MCMC candidate samples was seen to have a positive but very weak effect on final surrogate precision, suggesting that the runtime of MCMC on each iteration can be limited for increased efficiency. The most complex dynamics arose with the adjustment of sample increment, shown in~\cref{fig:qassincr}. For each tested initial sample quantity N, the optimal number of step samples was seen to be well-approximated by $\sqrt{N}$. The plotted error trends suggest that
incremental samples larger than this optimum give slower model improvement on both the training and evaluation sets, and a larger minimum error on the evaluation set. This performance distinction is predicted to be even more significant when trained on the expensive MC model, where the number of sample evaluations will serve as the primary bottleneck for computation time.
\begin{figure}[h!]
    \centering
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=1.1\linewidth]{fig6a_qassincrsamp.pdf}
    \end{subfigure}%
    \hfill
    \begin{subfigure}[t]{0.5\textwidth}
        \centering
        \includegraphics[width=1.1\linewidth]{fig6b_qassincrtime.pdf}
    \end{subfigure}
    \caption{QASS absolute training error over total sample quantity (left) and number of iterations (right). MAE represents surrogate error on the adaptively-sampled training/test set, and E\_MAE on the independent evaluation sets.}
    \label{fig:qassincr}
\end{figure}

@@

\begin{figure}[h]
  \centering
    \includegraphics[width=0.78\linewidth]{fig8_qasssampling100k.pdf}
    \caption{Absolute training error for QASS and baseline scheme, with 100k initial samples.}
  \label{fig:qasssampling100k}
\end{figure}

Such tests revealed that while QASS has unmatched performance on its own
adaptively-sampled training set, it is outperformed by the baseline scheme on
uniformly-random evaluation sets. We suspected that while QASS excels in
learning the most strongly peaked regions of the TBR theory, this comes at the
expense of precision in broader, smoother regions where uniformly random
sampling suffices. Therefore a mixed scheme was implemented, with half MCMC
samples and half uniformly random samples incremented on each iteration, which
is also shown in~\cref{fig:qasssampling}. An increase in initial sample size was observed to also resolve precision in these smooth regions of the toy theory, as the initial samples were obtained from a uniform random distribution. As shown in ~\cref{fig:qasssampling100k}, with~\num{100000} initial samples it was possible to obtain a ${\sim}40\%$ decrease in error as compared to the baseline scheme. Comparing at the point of termination for QASS, this corresponds to a ${\sim}6\%$ decrease in the number of total samples needed to train a surrogate with the same error.
